Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-05-04 Paper_Draft2/Draft2.tex
merge conflict resolved
delete citation

The AIF360 toolkit contains various group and individual fairness
metrics as well as pre-processing, in-processing, and post-processing
algorithms that we used to debias the COMPAS algorithm
\citep{aif360-oct-2018}. Fairness metrics are mathematical measures of
whether an algorithm treats members of different groups (such as racial
or gender groups) equally. An algorithm has no understanding of the
historical oppression of certain groups and how such bias is baked into
the data on which the algorithm is trained. Thus, fairness metrics
provide a method of evaluating an algorithm's level of bias towards or
against the unprivileged group versus the privileged group. We
researched the definitions and applications of different fairness
metrics \citep{ashokan2021fairness} to determine which metric would be
most appropriate for our project.

@@

Pre-processing refers to mitigating bias within the training data, and
it is the most flexible method because it has not yet trained a model
that may carry assumptions about the data. It is important to keep in
mind that pre-processing prevents assumptions in the modeling, but does
not account for the bias in data collection. Training data is where bias
is most likely to be introduced. We use the reweighing pre-processing
algorithm from AIF360 which assigns weights to the data. ``The advantage
of this approach is, instead of modifying the labels, it assigns
different weights to the examples based upon their categories of
protected attribute and outcome such that bias is removed from the
training dataset. The weights are based on frequency counts. However as
this technique is designed to work only with classifiers that can handle
row-level weights, this may limit your modeling options''
\citep{Ronaghan2019AI}. We used a logistic regression model for this
algorithm, as it is the easiest to interpret in the given context. After
running the fairness metrics using the pre-processing algorithm, we were
able to compare our results to the baseline metrics from the previous
section.

>>>>

The AIF360 toolkit contains various group and individual fairness
metrics as well as pre-processing, in-processing, and post-processing
algorithms that we used to debias the COMPAS algorithm
\citep{aif360-oct-2018}. Fairness metrics are mathematical measures of
whether an algorithm treats members of different groups (such as racial
or gender groups) equally. An algorithm has no understanding of the
historical oppression of certain groups and how such bias is baked into
the data on which the algorithm is trained. Thus, fairness metrics
provide a method of evaluating an algorithm's level of bias towards or
against the unprivileged group versus the privileged group. We
researched the definitions and applications of different fairness
metrics \citep{ashokan2021fairness} to determine which metric would be
most appropriate for our project. To begin, we implemented a logistic
regression model predicting recidivism using the defendant data. This
analyzes the baseline values without any bias mitigation. We use
logistic regression, as it is the easiest model to interpret in the
given context.

@@

Pre-processing refers to mitigating bias within the training data, and
it is the most flexible method because it has not yet trained a model
that may carry assumptions about the data. It is important to keep in
mind that pre-processing prevents assumptions in the modeling, but does
not account for the bias in data collection. Training data is where bias
is most likely to be introduced. We use the reweighing pre-processing
algorithm from AIF360 which assigns weights to the data. ``The advantage
of this approach is, instead of modifying the labels, it assigns
different weights to the examples based upon their categories of
protected attribute and outcome such that bias is removed from the
training dataset. The weights are based on frequency counts. However as
this technique is designed to work only with classifiers that can handle
row-level weights, this may limit your modeling options''
\citep{Ronaghan2019AI}. After running the fairness metrics using the
pre-processing algorithm, we were able to compare our results to the
baseline metrics from the previous section.
